\chapter{Expert Statements\ifdraft{ (Dan/Joe K.) (0\%)}{}}
\label{appendix:expert_statements}

The following expert statements are all included unedited, in their
entirety.

\section{Josh Beneloh}

\subsection*{The Viability of Responsible Internet Voting}

Remote voting\footnote{Remote voting is defined here as voting without
  the benefit of the public monitoring that takes place in a
  traditional poll site.}  entails significant risks above and beyond
those of in-person poll-site voting.  Included among these are risks
to integrity – as remotely-cast ballots may pass through numerous
hands without independent observation – and risks to privacy – as
voting takes place without the benefit of publicly-enforced voter
isolation.

Internet voting substantially exacerbates the risks of remote voting
by making it possible for small problems to be magnified and
replicated on a large scale.  Careless or malicious errors, intrusive
malware, and unforeseen omissions – all of which can be caused by
individuals or very small groups – can cause very large numbers of
votes to be changed and the privacy of large numbers of voters to be
compromised.

The technology known as end-to-end (E2E) verifiability allows
individual voters to verify that their intended votes have been
properly recorded and that all recorded votes have been properly
counted.  When applied to in-person voting, E2E-verifiability provides
new assurances to voters by allowing them to check for themselves that
the results of an election are correct.  When applied to Internet
voting, E2E-verifiability mitigates some of the risks described above
– but does not eliminate them:  voters are able to check that their
ballots are properly recorded and counted, but malware can still
compromise privacy, prevent voters from casting their ballots, and
otherwise hinder voters.

Although E2E-verifiable election technologies have existed for more
than thirty years, their use has thus far been limited to small
demonstration systems and private elections for student governments,
professional societies, and the like.  E2E-verifiable elections
produce new challenges and complications for implementers and
administrators.  They represent a new and different paradigm for
elections – substantially replacing the notion of verification of
election equipment with that of verification of the integrity of
individual elections.  As such, it is important to act deliberately
and gain experience with E2E-verifiability in more manageable
environments before attempting to deploy E2E-verifiable elections in
their most challenging environment:  the Internet.

These realities lead us to two principal conclusions.

\begin{itemize}
\item Public elections should not be conducted over the Internet
using systems that are not end-to-end verifiable.

\item End-to-end verifiable Internet voting systems should not be
used before end-to-end verifiable poll-site voting systems have been
widely-deployed and experience has been gained from their use.
\end{itemize}

The second of these two principles is also necessitated by the fact
that an E2E-verifiable election must have a tally to verify, and if an
E2E-verifiable system is used only for remote voters, then the votes
of these remote voters must be separately tallied and reported.  Few
jurisdictions are willing to segregate and report the tallies of local
and remote voters separately. 

We take no position here as to whether the integrity benefits of
E2E-verifiability and the privacy benefits it makes possible outweigh
the risks of remotely-executed large-scale corruption of an
Internet-based election, but we are agreed upon the conclusions that
“naked” Internet voting is dangerous and irresponsible and that
E2E-verifiability should be deployed in the less risky and more
manageable scenario of in-person poll-site voting before it is
deployed in the wilds of the Internet.

\section{David Jefferson and Concurring Experts}
{\Large\sffamily\bfseries\textcolor{DarkGreen}{\MakeUppercase{Candice
      Hoke, Ronald L. Rivest, Barbara Simons, Philip Stark, and
      Vanessa Teague, Concurring}}}

\subsection{Election security is national security}

In a democracy election security is a key part of national
security. The very legitimacy of government depends on the fact, and
also the public perception, that the outcome of every election fairly
represents the will of the people. We must be assured that no part of
the voting process is unfairly manipulated to produce a different
outcome, that all and only those who are eligible to vote have the
opportunity to do so, that no one votes more than once, and that the
privacy of the ballot is not compromised.

In this paper we demonstrate that while end-to-end verifiable (E2E-V)
voting systems have a great deal to offer, but when they are embedded
in an Internet voting context (E2E-VIV) they still do not provide
sufficient security to prevent remote attackers from silently
modifying votes and changing the outcome of elections undetectably, or
from disrupting the election and disenfranchising a large number of
voters. Fundamental security problems remain with E2E-VIV systems for
which there are no practical solutions in sight and that will not be
resolved in the foreseeable future.

\subsection{Verifiability}

Election security depends on \emph{verifiability}. After an election
is closed there must remain enough evidence for anyone who doubts the
results to re-examine and rationally determine whether the winners
were called correctly. We need to verify that only eligible voters
voted, that no votes were lost, or duplicated, or modified, that no
phony votes were inserted, that every eligible voter who tried to vote
was able to do so, and that all the votes were counted correctly, once
and only once. That evidence trail has to be \emph{end to end},
spanning the entire data path through which the ballot data travels,
from the voters' heads to the final result. We must be able to
reconstruct the vote totals (or statistically audit them) from the
original unmodified ballots or copies of those ballots that are
provably identical to the originals as intended by the voter.

The traditional approaches to verifiability are all based on
physically secured and indelible paper ballots (or paper cast vote
records) that can be recounted or audited by humans without having to
trust any software or complex machines. The goal of end-to-end
verifiable (E2E-V) systems is to use cryptographic protocols to
achieve for all-electronic voting systems the same (or higher) level
of confidence in the election outcome for electronic voting systems as
is achievable with paper-based systems. The goal of end-to-end
verifiable \emph{Internet voting} (E2E-VIV) systems is the same, but
specifically extended to the much more difficult security context of
remote voting from private platforms and devices over the public
Internet.

Some approaches to end-to-end ``verification'' at first sound great,
but fall far short. It is not sufficient, for example, to provide a
feature whereby each voter can verify that her own ballot was
correctly transmitted to the server over the Internet. First, it is
difficult to provide verification capability without also making it
possible for the voter to prove to a third party how she voted,
thereby enabling automated vote selling and voter coercion. But even
if that problem were resolved and if every voter verified that her
voted ballot was properly received, it would still not be possible to
demonstrate that no phony votes, unassociated with any actual voter,
were inserted.

\subsection{The power of E2E-V systems}

End-to-end verifiable voting systems are a major conceptual and
mathematical step forward from conventional voting systems. Through
advanced cryptographic techniques these varied systems, while
differing in many ways in their architecture and in the roles of
insiders, share the following fundamental security properties:

\begin{enumerate}[label={\alph*})]
\item \textbf{INTEGRITY.} Once a voter successfully enters her ballot
  into the E2E-V system it cannot be undetectably lost or modified in
  any way, even in the presence of bugs or malicious logic.

\item \textbf{PRIVACY.} Once a ballot enters the E2E-V system it is
  encrypted, so that there is no way that the privacy of the ballot to
  be violated subsequently.

\item \textbf{COUNTING ACCURACY.} The ballots cannot be miscounted
  without that fact being detectable.

\item \textbf{UNIVERSAL PUBLIC VERIFIABILITY.} The systems output and
  publish sufficient verification data so that \emph{anyone} can
  verify that no ballots were lost or modified and that the votes were
  properly counted. The verification data provides essentially a
  \emph{cryptographic proof} that ballot integrity was preserved and
  the counts are correct. Anyone is free to run a verification program
  over the verification data to confirm it. You don't even have to
  trust the official verification program---you can use one from a
  source you trust, or if you have the skill you can write your own.

\item \textbf{OPENNESS AND TRANSPARENCY.} The code for E2E-V systems
  is generally open source. The mathematical principles underlying the
  E2E-V security guarantees have been vetted by many cryptographic
  experts and are open and public. And the specifications for proof
  checkers are also publically documented so that mutually suspicious
  political groups can hire their own experts whose independent
  election verification programs, if correct, must agree.
\end{enumerate}

These powerful E2E-V security properties are not shared by any
traditional voting system. In a precinct voting context they make
E2E-V systems essentially invulnerable to ordinary software bugs, to
malicious code inside (but not outside) the voting system, to
transient hardware faults, and to most kinds of insider fraud (at
least without a large conspiracy). Such failures will at least be
detected because any lost or modified ballots and any miscounts will
be flagged whenever anyone runs a verification program. This is in
strong contrast to other forms of purely electronic voting systems
which are unverifiable, and in which bugs or malicious logic can cause
errors that are totally undetectable.  The wrong people may wind up
taking office without anyone knowing that the election results were
incorrect.

For these reasons it is appropriate to consider E2E-V systems in any
electronic voting situation. E2E-V adds truly powerful security
guarantees, particularly in a precinct voting situation where voter
authentication is done in person and where we have good reason to
presume that the certified software in the voting machines is not
malicious.

But as we explain in the next section, these E2E-V security guarantees
\emph{do not fully extend to Internet voting systems}. E2E-V Internet
voting systems (E2E-VIV) have exploitable security holes for which
there are no good solutions today and that preclude them from being
suitable for use in public elections for the foreseeable future.
  
\subsection{Remaining unsolved security issues with E2E-VIV systems}

Once the voters' choices are safely input to a precinct-based E2E-V
system many, but not all, of the security guarantees described in the
last section follow directly. But that is a key qualification: these
guarantees begin \emph{once the voters' choices are safely input to
  the E2E-V system}, but not before. Unfortunately, when an E2E-V
system is embedded into the Internet voting context as an E2E-VIV
system, new security problems appear that the E2E-V guarantees do not
address and that cannot, with any current technology, be fixed. The
problems with E2E-VIV systems arise \emph{before the votes even enter
  the system}. In this section we enumerate the remaining difficult
security problems that will have to be solved definitively before we
can consider deploying an E2E-VIV system.


\subsubsection{Voter authentication}

A central issue with all remote, online voting systems is voter
authentication. The voting system must be able to positively identify
the voter in strong a way, so that it is essentially impossible to
avoid the authentication process, and impossible to fool it so that an
ineligible person is allowed to vote or that someone can fraudulently
impersonate another voter.

Voter authentication is not part of an E2E-VIV system, but is a
separate security issue. Unfortunately it is a very difficult and
complex problem that remains unresolved (in the U.S. at least) for the
foreseeable future.

Strong voter authentication is required for several reasons. Any
online voting system must:

\begin{itemize}
\item verify that potential voters are duly registered or eligible to
  vote in the jurisdiction they attempt to vote in;
\item prevent anyone from voting more than once; and
\item resist vote selling, vote coercion, and proxy voting insofar as
  possible in a remote voting situation.
\end{itemize}

In an online voting system it is not sufficient to use the kinds of
authentication commonly used in ecommerce situations, e.g. passwords,
challenge-response systems based on personal information, or
email-confirmations. These very weak authentication systems are more
or less sufficient in commercial situations where secrecy is not so
important, and where fraudulent transactions can be detected
eventually and frequently be reversed or at least can be absorbed as a
cost of doing business. 

But in the national security context of an online election such weak
authentication mechanisms will not suffice. If an attacker has the
technical means to impersonate one voter, he can generally automate
and amplify his methods to impersonate thousands of voters with very
little additional effort. Almost every month we hear of huge data
breaches at commercial or government institutions that have already
allowed vast amounts of personal information on tens of millions of
people to fall into the hands of criminals or foreign powers. Thus,
any authentication mechanisms based on merely presenting personal
information (name, address, account number, driver’s license or social
security number, mother’s maiden name, etc.) is hopelessly compromised
already, and way too weak for use in an election. Unfortunately some
states have implemented such embarrassingly weak online authentication
systems and have been forced to strengthen them, though they are still
not sufficiently strong.

The traditional voter authentication method is based on wet ink
signature matching. The voter is required, either in person at the
precinct or on the envelope of a mail-in ballot, to duplicate with a
new ink signature the old signature image on file from the time she
registered to vote. In some states this is augmented with VoterID
requirements at the polls. But there is no way to securely
(unforgeably) input a wet ink signature image to a computer or mobile
device and transmit it over the Internet for authentication with the
ballot. Nor is there yet a way to securely and unforgeably transmit
any of the usual VoterID documents.

Many people have suggested voter authentication systems based on
biometrics such as fingerprints or retinal scans. For very good
reasons too numerous to fully explain here none of these mechanisms is
suitable for online voting. And it is important to note that while
some mobile phones and tablets have fingerprint authentication devices
built in, such systems authenticate the user to the device only. They
do not authenticate the user to any remote service over the Internet,
nor can they easily be extended to do so securely.

Other stronger, more technical authentication methods could be
considered.  Voters could be issued cryptographic ID cards such as the
CAC cards issued to DoD personnel or like the national ID card of
Estonia. Cryptographic ID cards would in principle enable voter
authentication from any Internet-connected computer or device that
could read them. But no U.S. state issues such IDs to its citizens or
voters, and it seems unlikely that any will do so in the foreseeable
future. Even if the security climate changes and people are willing to
accept such an ID system, the startup and maintenance costs will be
very high.  Voters would have to buy computers or devices that could
read the cards, and they would almost certainly have to be useful for
other online purposes besides just voting in order to justify the
costs involved to both the government and the voter.

The fact is that the U.S. has no strong, universally deployed online
citizen identification and authentication system, and none is on the
horizon. While strong remote voter authentication is not a
fundamentally unsolvable problem, it is an immense practical problem
that has to be dealt with before we can consider deploying any online
voting system, including E2E-VIV systems.

\subsubsection{Client side malware}

In an E2E-VIV system voters compose and input their vote choices on a
privately owned (hence unsecured) platform, either a PC or mobile
device. If the voting platform is infected with malware or spyware, it
is complicated to prevent the votes from being modified or reported to
a third party, and impossible to prevent them from just being thrown
away by the malware before the ballot is encrypted and enters the
E2E-VIV system.

The malware threat is ubiquitous now and is fundamental to all online
voting systems. No device is safe from malware infection. No software
safeguards such as commercial antivirus systems are very effective
against it. There are hundreds of ways that malware can infect a
voting platform, sometimes by sophisticated technical means and
sometimes by tricking users into taking unsafe actions. There are
hundreds of places in the huge multilayered software ecosystem of a PC
or mobile device where malware can hide and be launched from. And
there are thousands of hardware, OS, browser, combinations, and
countless configuration choices---way too many for any possible
comprehensive defense against malware. A modern PC or mobile device
can easily contain software elements from a hundred different
companies or open source development groups, any one of which may
either be malicious or contain critical vulnerabilities that enables
malicious code. Such vulnerabilities are so numerous that more are
discovered all the time and vendors release security updates on a
regular basis to plug the more recently discovered holes. E2E-VIV
systems have no ability to prevent or even detect the actions of
malware before the votes are safely submitted into the E2E-VIV
software.

Malware in voters' computers can undermine the election in three
fundamental ways. 

\begin{enumerate}[label={\roman*})]
\item \textbf{MALWARE MODIFICATION OF VOTES.} Malware may actually
  modify the voter’s choices surreptitiously, before they are
  submitted to the E2E-VIV system. The techniques for accomplishing
  this without tipping off the voter will depend on the detailed
  architecture of the voting system, e.g. whether the client side is
  packaged as a full-blown application, or as a mobile app, or a
  Javascript script, or a browser plugin, or some other form. But in
  all cases the voter’s choices must be input to the PC or mobile
  device in the clear, unencrypted, and be processed by a large amount
  of system software and application/browser/script software
  \emph{before} it is encrypted and submitted to the E2E-VIV
  system. Regardless of the E2E-VIV architecture, with today’s
  software tools it is reasonably straightforward for malicious code
  to modify votes undetectably before they are encrypted.

  Depending on the design of the E2E-VIV system, it may be possible
  for some voters to discover after the fact that different votes were
  recorded for them than the ones they thought they cast. But even so,
  there will generally be no way to prove to election officials that
  the voter did not cast the votes recorded for her by mistake, or
  cast them deliberately and then change her mind. Whether there is a
  remedy for voters who claim their votes are modified by malware and
  falsely recorded is an unresolved question.

  We cannot generally eradicate the threat of malware. But in the
  special case of online voting there are techniques that in theory
  can prevent malware from surreptitiously modifying
  votes. Unfortunately they all involve additional burdens on the
  voter in the form of code voting, or special hardware devices
  independent of the PC, or a second independent communication channel
  to the election server that does not use the Internet, or at least
  is guaranteed to use an independent path from the one the votes
  travel. All known methods of working around client side malware
  involve some complication in the voting process that will be enough
  of a barrier, at least for the time being, to discourage many
  voters.

\item \textbf{MALWARE VOTE PRIVACY VIOLATION.} Malware on a voter’s
  computer or mobile device could allow her completed ballot to enter
  the E2EVIV system unmodified, but prior to that it could \emph{also}
  send a copy of her votes to a third party. Unless a voter has
  considerable expertise and has special instrumentation running
  during the voting transaction there is no way for her to know
  it. And if the instrumentation was not in place before voting there
  is no after-the-fact test that can determine whether this happened,
  and certainly no way to reverse the privacy violation. If the voter
  is voting from a mobile device, as opposed to a PC, often no such
  instrumentation even exists today.

  In any remote voting situation there is always the possibility that
  someone can physically look over a voter’s shoulder and watch her
  vote. That is a risk we live with also with paper mail-in
  ballots. But the main concern is not with individual cases of
  privacy violation, but with widespread automated spying on many
  online votes.

  Widespread vote privacy violation can undermine democracy in two
  major ways. First, in situations where some people have power over
  others (e.g. employers, commanding officers, union supervisors,
  parents, nursing home management) revealing who cast which ballot
  can be the basis for coercion or retaliation. This may not (yet) be
  a widespread concern in the U.S., but it certainly is in other
  countries.

  Also, automated privacy violation can enable large scale, automated
  vote buying and selling.  It is easy to imagine a scheme in which
  many voters are induced to sell their voting credentials, or to run
  a particular program while voting for a particular candidate in
  exchange for PayPal dollars or some other crypto currency such as
  Bitcoin that can be transmitted entirely online. The vote buying
  transaction would likely be totally undetectable by
  authorities. Even if the scheme eventually comes to the attention of
  authorities, the buyers may be long gone, or may be on foreign soil
  out of reach of U.S. law.  In any case there would be no way to know
  how many votes were sold or who the sellers are. Technical tricks,
  such as allowing voters to vote multiple times online with only the
  last cast vote actually counting, are not effective when the
  attacker knows how the system works or when the voter cooperates in
  a vote sale.

  As with malware that intends to modify ballots surreptitiously,
  there are workarounds that can prevent malware from surreptitiously
  revealing how someone votes to a third party. But again, they
  complicate the process of online voting sufficiently to be a barrier
  that will discourage many voters from voting

\item \textbf{MALWARE DENIAL OF SERVICE.} The easiest and most
  intractable malware attack is one that simply prevents the voter
  from successfully voting. That can be done in many ways. The malware
  could make it appear that there was an error of some kind, which
  might be frustrating but hardly surprising to voters who might
  either blame themselves or their own flaky computers or attribute it
  to just another buggy online service. Alternatively, the malware
  might perfectly mimic a completed voting transaction, so that the
  voter believes she has successfully voted, whereas the malware would
  simply throw the ballot away.

  Such a denial of service attack might not be very politically
  effective if it is applied to a random set of voters.  But if it can
  be applied \emph{selectively} to voters who would be likely to vote
  in a way that the attacker does not like, then it becomes a powerful
  partisan fraud tool.  The malware writer may want to make a good
  guess as to how the voter will likely vote before deciding whether
  of not to block her from voting.  Fortunately, there are many clues
  in a voter’s computer or mobile device, such as browser history, to
  indicate at least a likely party preference or social class, and
  that is probably all the information the malware would need.

  Some voters may be sophisticated enough to detect that their ballots
  were never included in the count, especially during the
  post-election verification stage when some might discover that there
  is no record that they voted. But any particular voter would find it
  almost impossible to prove to election officials that she actually
  tried to vote online but that malware prevented it. Perhaps she
  simply never really tried to vote, or for some other technical
  reason not related to malware she had been prevented from
  successfully voting. There would be no evidence anywhere accessible
  to officials that could help them diagnose the situation. Even if
  the voter brought her computer in for forensic examination by
  experts, chances are that the malware would have erased evidence and
  erased itself, leaving no trace.

  And finally, even if an obvious widespread malware denial of service
  attack were somehow discovered, there would be no way to estimate
  how many ballots were lost and how many voters were
  disenfranchised. The E2E-VIV system does nothing to help with such
  an estimate because the ballots are discarded by the malware before
  they ever enter the E2E-VIV system.

  Unfortunately, while there are (at best inconvenient) workarounds
  for malware that aims to surreptitiously modify a ballot or send it
  to a third party, \emph{there is fundamentally no workaround for
    malware designed to just prevent voting}.  Well-designed malware
  would make the voter believe she had successfully voted, and she
  might never discover until it was too late that she did not. Even if
  she did discover it, the only recourse would be to vote from a
  different, uninfected PC or device, \emph{but she almost certainly
    would not know that malware was the cause of the problem and would
    likely not know to vote from a different machine!}

\end{enumerate}

Malware is a profound, absolutely fundamental problem that has been
with us since the dawn of the PC age or before and will be with us for
as far into the future as we can see. There is fundamentally no way to
totally eradicate client side malware, or totally immunize against it,
or even detect its presence. Malware is getting ever easier to write
because templates, kits, libraries, and exemplars of successful
malware are widely available to aid attackers, and because the payoff
far exceeds the risks of getting caught. It is estimated that anywhere
from 10 to 30 percent of all PCs in the world are infected with
malware, and even more when spyware is included. And there are
probably no reliable estimates as to the fraction of mobile devices
similarly infected.

Finally, even if an easy to use, accessible workaround for client side
malware is invented that preserves vote integrity and privacy, it will
still not be possible to prevent a malware denial of service attack
that just blocks voting. Such denial of service attacks are in a
theoretically different category and there will never be a general way
to thwart them all, nor a general way for voters or election officials
to unambiguously recognize one. Even if it is recognized, there will
be no way to estimate how many voters were affected.

The conclusion therefore, is that client side malware remains a
fundamental threat that E2E-VIV systems cannot fully defend against. 

\subsubsection{Network attacks and Distributed Denial of Service
  (DDoS) attacks}

E2E-VIV systems are client-server protocols that execute on top of
several layers of other software, including the operating system and
browser on the client side, the operating system and server complex on
the server side, and the various levels of TCP/IP protocol stack all
through the Internet, as well as routing protocols, DNS, NTP, DHCP,
and also numerous other protocols used in wireless or mobile
devices. The E2E-VIV system cannot work properly unless all of this
other software works properly also. We have already discussed the
problem of malicious logic on the client side. But E2E-VIV software is
also attackable from the server side or from the Internet
infrastructure software that the E2E-VIV software depends on.

There are many ways to attack an E2E-VIV election by maliciously
modifying or configuring the software in the Internet. Such attacks
are called \emph{network attacks}. Any IT person who controls a
router, DNS server, or another element of Internet infrastructure is
in a position to prevent votes from getting to their destinations.  On
the positive side, E2E-VIV systems are partly robust against such
network attacks in that they cannot result in votes being falsely
injected or modified without detection. This is a clear advantage that
E2E-VIV systems have over other Internet voting systems. However,
there is no way to prevent a network attack from disrupting the
E2E-VIV protocols in a way that causes ballots to be lost,
i.e. undelivered.  While this will also be detectable, the malicious
loss of votes in transit cannot be prevented by an E2E-VIV protocol,
and it may not be possible even to estimate the number of votes
affected.

One especially dangerous form of network attack is the
\emph{distributed denial of service} (DDoS) attack.  In this attack,
an attacker floods the server (or some other subsystem) with so much
traffic or other work that it either crashes the system or else slows
it to such a crawl that it is effectively down. Voters would
experience a DDoS attack on a vote server as either
\emph{extreeeeemly} long waits between steps in the voting process, or
total nonresponsiveness of the server, or some other error. The net
effect is that large numbers of voters would simply be
disenfranchised. The attack can be directed pointedly at the server
side, in which case all online voters would be affected, or it could
be selectively directed at certain parts of the Internet
infrastructure that would affect only a subset of the voters.

We have to be able to defend online elections against DDoS attacks for
two key reasons.  First, they are about the easiest of all network
attacks to perpetrate.  There are many different kinds of DDoS attack
against different parts of Internet infrastructure and different
levels of software, and there are many kits available on the dark net
to allow anyone from anywhere in the world to perpetrate a DDoS attack
against almost any target. In fact, the means of DDoS attack are so
routinized and ubiquitous that there are illegal businesses online
that will conduct an attack to your specifications against any target
you choose for a moderate price. You can ask for, say, a 50 gigabit
per second attack for the last 4 hours on Election Day against the IP
address of the (hypothetical) Cook County vote server. That would
probably prevent anyone from voting online in that jurisdiction during
those hours. All of those voters would be disenfranchised, but none of
them would be able to prove that they were among of the victims, and
election officials would not even be able to make a decent estimate of
the number of ballots lost.

DDoS attacks have actually been used in real public elections around
the world at least four separate times that have been made
public. (Arizona Democratic Primary, 2000; Ontario NDP, 2003; Hong
Kong people's election, 2012; NDP of Canada 2012). While there are
various tools that can be used to \emph{ameliorate} some DDoS attacks,
there is no \emph{general solution}, and the DDoS problem is so
fundamental that there will probably never be one with the current
architecture of the Internet. Vulnerability to DDoS attacks is
effectively built in to its design. Hence, all E2E-VIV systems are
vulnerable to network attacks that can result in disenfranchising a
large number voters with no way of even measuring how many were
affected. There is no fundamental defense against DDoS attacks.

\subsection{Conclusion}

E2E-V offers a dramatic improvement in the security of voting
systems. While it is necessary for \emph{any} online voting system for
public elections, it is by no means sufficient. Once it is embedded in
a larger \emph{Internet voting} context fundamental new security
vulnerabilities appear for which there are no solutions today, and no
prospect of solutions in the foreseeable future. These include
vulnerability to authentication attacks, client side malware attacks,
and DDoS attacks that can be perpetrated by anyone in the
world. Unless and until those additional security problems are
satisfactorily and simultaneously solved---and they may never be---we
must not consider any Internet voting system for use in public
elections.
