\chapter{Remote Voting\ifdraft{ (Philip) (100\%)}{}}
\label{chapter:remote_voting}

\section{Rationale}
Remote voting is becoming increasingly common, prompted by the growing
and diverse needs of voters. Remote voting is used to enable overseas
citizens and military personnel to participate in elections, reduce
access related discrimination domestically, and decrease expensive
administrative overhead of polling locations.  In the United States,
fewer than 5\% of ballots cast in general elections during the 1980s
were cast before election day. By the 2012 general election, 31\% of
all ballots were cast early, and 17\% were cast by mail. The states of
Washington, Oregon, and most recently Colorado have entirely switched
over to all-mail voting. For an election system fully enfranchise the
electorate, it must treat remote voting as a first-class capability
rather than as a backup system with second-class effectiveness, speed,
security, and integrity.

\subsection{Accessibility}
According to the International Center for Disability Information and
the National Institute on Disability and Rehabilitation Research, 20\%
of Americans live with disabilities. The Voting Accessibility for the
Elderly and Handycapped Act of 1984 mandates that any person with a
disability may vote remotely without having to present medical
documentation, reducing the barriers to remote voting for those with
accessibility needs.

%The Help America Vote Act (HAVA) of 2002 requires that all polling places in elections for federal office, anywhere in the United States have at least one voting system  \ldots
%\ldots 20\% of U.S. adults with disabilities say they have been unable to vote in presidential or congressional elections due to barriers at or getting to the polls.

\subsection{UOCAVA}
In 1986, Congress enacted the Uniformed and Overseas Citizens Absentee
Voting Act (UOCAVA) to address the needs of citizens in the uniformed
services, merchant marines, and other overseas civilians. UOCAVA
mandates that these voters be able to register and vote remotely in
federal elections. It is difficult to calculate the exact number of
UOCAVA eligible voters, but \autoref{fig:uocava_populations} lists a
recent estimate of the total.

\todoacf{Ask Philip for context on these numbers. Statewise breakdown
  doesn't seem to add much, and the difference between what the two
  are counting is unclear}

\begin{figure}
\begin{center}
\begin{tabular}{l p{.3\textwidth} p{.3\textwidth}} % this can be made prettier
  {\bf State} &
  {\bf Overseas Voting Eligible \newline Population} \newline(McDonald 2009) &
  {\bf Overseas military and federal \newline civilian employees}
  \newline(US Census Bureau 2010)\\\hline\\
Texas & 11.05\% & 11.78\%\\
California & 9.78\% & 8.44\% \\
Florida & 9.09\% & 9.54\% \\
New York & 5.31\% & 4.12\% \\
Pennsylvania & 4.10\% & 3.12\% \\
Illinois & 4.03\% & 3.24\% \\
Ohio & 3.51\% & 3.07\% \\
Michigan & 3.29\% & 2.68\% \\
Georgia & 2.84\% & 3.83\% \\
Washington & 2.78\% & 2.77\% \\
North Carolina & 2.78\% & 2.91\% \\
Tennessee & 2.57\% & 2.81\% \\
Virginia & 2.51\% & 3.52\% \\\hline\\
Estimated Total & 4,972,217 & 1,042,523
\end{tabular}
\end{center}
\caption{Comparisons of American Overseas Population by State}
\label{fig:uocava_populations} % change to table?
\end{figure}

\subsection{Domestic Absentee}
Domestic absentee voters are those who vote early in-person, or cast
their votes cast by mail because they are unable or do not want to be
present at polling locations on election day (this excludes UOCAVA
voters, and voters in states that vote exclusively by
mail). 21,853,762 or 16.6\% of all votes cast in the 2012 US general
elections were from domestic absentee voters \cite{eac2012survey}. As
of 2015, 27 states allow voters to apply for an absentee ballot
without providing a justification, known as ``no-excuse absentee
voting''.

\todoacf{Take out this figure too?}
% proportions of domestic absentee voters. These states are frequently
% using domestic absentee to address concerns and costs associated with
% administering polling locations in low population density areas.

\begin{figure}
\begin{center}
\begin{tabular}{l c}
{\bf State} & {\bf Percent of Population}\\\hline\\
Colorado & 71.4\%\\
Arizona & 65.9\%\\
Montana & 57.5\%\\
Georgia & 48.8\%\\
Iowa & 43.1\%\\
California & 39.8\%\\
Hawaii & 36\%\\
North Dakota & 28.8\%\\
Florida & 26.8\%\\
Michigan & 26.4\%\\
Wyoming & 26.2\%\\
Maine & 25.5\%\\
Nebraska & 25.4\%\\
Idaho & 24.3\%\\
Ohio & 22.4\%\\
Wisconsin & 21.4\%\\
Vermont & 20.4\%\\
\end{tabular}
\end{center}
\caption{Votes Cast as Domestic Absentee 2012 General Election}
\label{fig:domestic_populations}
\end{figure}

\subsection{Expectations}

In 1952, a study by the American Political Science Association defined
ten voting rights necessary for members of the armed
services~\cite{american1952findings}. Although initially defined for
military voters, these rights have served as the basis for defining
the expectations of all remote voters through UOCAVA and other
subsequent legislation. Among these rights are:

\begin{enumerate}
  \item To vote without registering in person.
  \item To vote without paying a poll tax or having to meet
    unreasonable requirements
  \item To use the Federal postcard application both to register and
    to request a ballot, rather than having to use state-specific
    paperwork.
  \item To receive ballots for primary and general elections in time to vote.
  \item To be protected in the free exercise of their voting rights.
  \item To receive essential information needed to vote.
\end{enumerate}

These rights first found their way into law in the Federal Voting
Assistance Act of 1955 (FVAA), but due to partisan struggles, the
rights were watered down from requirements into recommendations,
leaving much of the decisions about final implementation to the
states. Over time, subsequent legislation has strengthened these
recommendations into guarantees and requirements, and has expanded
rights to include participation by non-English speakers and people
with disabilities.

% \todoacf{See todo above; this seems out of place}

% The Help America Vote Act of 2002 (HAVA) defines
% additional mandatory minimum standards for states that specifically
% address voter access concerns raised in the 2000 general
% election. This act requires that elections support participation by
% non-English speakers and disabled persons by providing the same
% opportunities for access and participation, including voter privacy
% and independence. Additionally, HAVA mandates that voters whose
% eligibility is questioned must be permitted to cast a provisional
% ballot to be reviewed later by election officials.

% Finally it is most important to include the expectation that votes cast by
% registered voters are counted correctly while preserving privacy. As will
% be mentioned later in this chapter, there are many examples where this
% isn't the case.

% HAVA also put stricter auditing requirements in place, requiring
% voting technologies to produce a Verifiable Voter Paper Audit Trail
% (VVPAT); while preserving the privacy of the voter and the secrecy
% of the cast ballot.


\section{History}
\subsection{Armed Forces Voting}
Before the American Civil War, US citizens primarily voted in their
places of residence, and many states legally barred the casting of
votes from outside state borders. There was little effort from any
state to accommodate absentee voting. However, in 1864, with the Civil
War displacing soldiers from their residences, Lincoln's re-election
was at risk. With much lobbying on behalf of the Republican Party (and
opposition from the Democratic Party), nineteen Union states adopted
absentee voting procedures for military voters in time for the
election. Since the motivation for passing these laws was to secure
Lincoln's re-election rather than permanently expand voting access,
many absentee military voter laws were treated as temporary and
repealed after the war.

For the 1918 midterm elections, the US War Department decided that it
was not ready to support the military vote, going so far as to
prohibit individual states from canvassing overseas soldiers serving
in the First World War. The Second World War inspired another push for
the military vote in hopes of supporting the re-election of the
presidential incumbent. This prompted the Soldier Voting Act (1942)
which, although passed too late for the 1942 midterm elections, gave
military personnel absentee voting rights for federal elections during
times of war without subjugation to voting tax or postage
costs. Nonetheless, the act was notable for mandating that all
overseas voting would be regulated at the federal level and
implemented at the state level, a structure that continues to this
day. By 1944, partisan politics led to the weakening of the state
mandate from a requirement to a recommendation, leading to a 29.1\%
turnout rate vs. the 60\% domestic turnout rate~\cite{smith2015}.



\subsection{Remote Civilian Voting}

Progress for civilian absentee voters lagged behind progress for
military voters. In 1896, states began introducing civilian absentee
voting legislation. By 1924 only three states had no absentee voting
legislation, but state laws were a confusing and inconsistent
patchwork, lowering absentee turnout. Major progress for civilian
absentee voters would only come with legislation motivated primarily
by military voters such as the FVAA.

In the 1960s, lobbying from overseas civilian groups led to amendments
to the Voting Assistance Act expanding the number of civilians covered
by the law, though those amendments were once again voluntary
recommendations to the states. As lobbying pressure increased further,
the Overseas Citizens Voting Rights Act (OCVRA) passed in 1974 for the first
time guaranteed, rather than merely recommended, absentee voting
rights for overseas civilians.

In 1986, the Uniformed and Overseas Citizens Absentee Voting Act
(UOCAVA) was passed, combining and replacing the FVAA and the OCVRA
and making the rights recommended by the previous acts into
requirements for both military and overseas civilian voters.

\subsection{Disabled Civilian Voting}
The Voting Rights Act (VRA) of 1965 was the first legislation to
enfranchise voters with disabilities. The VRA granted voters who
require assistance to vote by reason of blindness, disability, or
inability to read or write assistance by a person of the voter's
choice.

The Voting Accessibility for the Elderly and Handicapped Act of 1984
(VAEHA) was passed to improve access for disabled and elderly
individuals. However, as with the recommendations of FVAA and OCVRA,
states were left to set their own standards of access, and limited the
disabled voters group to those with {\em physical disabilities}. The
VAEHA did, however, mandate ``no notarization of medical certification
shall be required of a voter with a disability with respect to an
absentee ballot or application for such ballot.''

The 1990 American Disabilities Act (ADA) requires that people with
disabilities have access to basic public services, including the right
to vote. The ADA does not strictly require that polling locations are
accessible. however it did extend the definition of disability to:
\begin{quote}
``a person who has a physical or mental impairment that substantially limits
one or more major life activities, a person who has a history or record of such
an impairment, or a person who is perceived by others as having such an
impairment.''
\end{quote}

The majority of federal laws passed to protect voting rights of
disabled citizens have struggled to clearly define a representative
range of disabilities, and are often focused on in-person access to
physical polling locations, which is expensive for states to
implement. Additionally, state-defined policies often ignore rights to
voting privacy, and exclude persons with multiple disabilities for
which assistive technologies are not yet available.

% Unfortunately most states claim to be unable to handle the costs
% associated with providing appropriate voting technologies.  much of
% this cost falls on the state to implement, and is co designate
% restrictions on physical polling locations.  With the passage of the
% American Disabilities Act (ADA) in Voting Accessibility for the
% Elderly and Handicapped Act of 1984 VVPAT
\subsection{Modern Remote Voting}

In 2002, Congress passed the Help America Vote Act (HAVA) in response
to problems found in gathering, counting, and auditing votes in the
2000 presidential election. HAVA requires that all polling places in
elections for federal office anywhere in the United States have at
least one voting system capable of assisting disabled voters,
addressing some accessibility concerns.

HAVA is also a response to the large number rejected ballots in the
2000 election and an inability to sufficiently audit ballots. HAVA
recommends election systems produce a Verifiable Voter Paper Audit
Trail (VVPAT) while preserving the privacy of the voter and the
secrecy of the cast ballot. HAVA also created the Federal Election
Assistance Commission (EAC) to oversee the development of new voting
machine standards, and released the Voluntary Voting System Guidelines
to aid in this transition.

The Military and Overseas Voter Empowerment (MOVE) Act of 2009
addresses barriers to overseas voter participation, specifically
attempting to reduce the number of ballots that were not counted due
to arriving late. The MOVE Act requires states to transmit absentee
ballots at least 45 days before election day, make all registration
material and blank ballots available electronically, and remove
notarisation requirements on all election material.

%\subsection{Integration with Local Elections}

%Every state has their own requirements, deadlines, and transmission
% restrictions which the FVAP documents in a 'Voting Assistance Guide'
% distributed to potential UOCAVA voters.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Historically, legislative progress on absentee voting moves
slowly. Existing voter rights regulations are enforced at a state
level and are hampered by local political attitudes. In 2010, the
Uniform Law Commission, a non-profit organization, oversaw drafting of
the Uniform Military Services and Overseas Civilian Absentee Voters
Act (UMOVA). UMOVA is designed to identify and standardize the
important protections and benefits found in federal legislation like
UOCAVA and MOVE in state and local elections. As of April 2015,
fourteen states and the District of Columbia have enacted UMOVA.

\section{Shortcomings of Current Practice}

Despite years of progressively stronger legislation addressing the
needs of remote voters, there are many shortcomings of current
election practices. The topics listed below draw from specific
concerns that have a large impact on remote voting participants.

\subsection{Use of Communication Technologies}
The majority of remote voting takes place via postal mail which is
subject to many inherent faults that are exacerbated for voters. The
2008 Post-Election UOCAVA Survey Report and Analysis found that 52\%
of attempted UOCAVA votes were not counted due to problems in the mail
delivery process. Additionally, maintaining correct voter registration
information for military voters and others who frequently move to new
addresses while abroad is expensive and prone to error.

% \begin{quote} {\em \ldots want to make comment about Florida, but concerned that it may be taken as political}\end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Entire process for UOCAVA voters generally can take up between 2 weeks and 2.5 months
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To address differences between states' absentee registration and
voting practices, the Federal Voting Assistance Program provides an
Voting Assistance Guide for registration forms. Unfortunately, this is
very large and difficult to follow. For states without streamlined
online registration, this has resulted in many failed voter
registration attempts.

\todoacf{doesn't the Federal Postcard Application address this?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{itemize}
%\item The major motivations for use of internet and communication technologies for UOCAVA has been to address ballot transit time, and simplify voter registration.
%\item OVF's streamlined website for FPCA in states that allow online registration
%\end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Accessibility and Usability}

In 2007, 20\% of Americans with disabilities said they were unable to
vote in presidential or congressional election due to barriers at or
getting to the polls~\cite{runyan2007improving}. This is frequently a
consequence of the voting technologies used and the physical location
of polling places. In the 2000 presidential election, 56\% of randomly
sampled polling places in the United States had at least one
identified impediments for disabled voters~\cite{united2001voters}.

For in-person voting, disabled persons often forfeit privacy in order
to receive assistance from an aide, often because polling locations
have of insufficient assistive technology, or because the assistive
technology in place is too difficult for both voters and poll workers
to use. Remote voting still presents obstacles: those with dexterity
impairments often have problems with handling and marking paper
absentee ballots.

\subsection{Auditing}

Although it is believed that voter fraud is fairly uncommon, it is a
major concern in a bipartisan system. It is very difficult to detect
voter fraud without depressing turnout or disenfranchising legitimate
voters. Policies intended to reduce fraud or protect identities, such
as increasingly prevalent and strict voter ID requirements, often
result in a higher rate of rejected ballots even for remote voters. In
the 2012 general election, over 20\% of absentee ballots were rejected
due to non-matching signatures or unmet ID
requirements~\cite{eac2012survey}.

% Ten states explicitly require a privacy waver if a voter uses fax or e-mail to return a voted ballot''

\subsection{Voter Privacy}

Voter privacy is a key expectation of a fair voting system. Privacy
promotes voter independence, and helps prevent voter coercion and vote
buying. Most remote voting practices require that voters forfeit
independence or privacy due to the physical reality that election
officials can enforce privacy in polling places, but not in voters'
living rooms. Several states even require signing a voter privacy
waiver when casting a remote ballot~\cite{smithtime}.

% In several jurisdictions, UOCAVA votes are
% not counted until it is determined that they may sway the election.
% perhaps nice to have cost analysis

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsubsection{Digital vs. Physical}
%\subsubsection{Risk-Limiting Audits}
%Risk limiting audits use a public random auditing process to make a strong
%argument about the statistical confidence of a particular election result. 
